---
title: "CM515 Day 1: Plotting with ggplot()"
output: pdf_document
---

```{r setup, include=FALSE}
library(datasets)
library(tidyverse)
library(knitr)
library(ggplot2)
```

## Goal: Utilize tidy data to generate complex graphs with few lines of code

## Start by loading the data

```{r load the data}
mpg <- mpg
```

## Check out the data

```{r check out the data}
# What are 5 functions we could use to explore the mpg dataset?


# Which manufacturer has the most models in this dataset?

```

This dataset suggests many interesting questions. How are engine size and fuel economy related? Do certain manufacturers care more about fuel economy than others? Has fuel economy improved in the last ten years? We will try to answer some of these questions, and in the process learn how to create some basic plots with ggplot2.

### Every ggplot has three key components:

-   Data
-   Aesthetic mappings between variables in the data
-   A layer to render the information (geom)

### A simple example

```{r simple example}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()

# Fill in the following:
# Data:
# Aesthetic: 
# Layers: 
# Plus sign location: 
```

### x and y aesthetic can be implied

```{r implied aesthetic}
ggplot(mpg, aes(displ, hwy)) +
  geom_point()

# What conclusions can you make from this graph?
```

```{r comparing plots}
# Do these data look similar when comparing engine displacement to city miles per gallon?
ggplot(mpg, aes(displ, )) +
  geom_point()

# We can more easily compare the highway and city miles per gallon using the gridExtra package.
library(gridExtra)

p1 <- ggplot(mpg, aes(displ, hwy)) +
  geom_point()

p2 <- ggplot(mpg, aes(displ, )) +
  geom_point()

grid.arrange(p1, p2, ncol = 2)
```

## Aesthetic attributes

### Color

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = "blue"))

# Value "blue" is scaled to red color
```

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(color = "blue")

# Points are given color
```

```{r}
ggplot(mpg, aes(displ, hwy, color = class)) + 
  geom_point()

# Unique color based on class of vehicle

# What additional information does the color add? What patterns do you see?
```

### Facet wrap

```{r facet wrap}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  facet_wrap(~ class)

# Read the documentation for facet_wrap. What arguments can you use to control how many rows and columns appear in the output?


#What does the scales argument to facet_wrap() do? When might you use it?


```

### Geoms

#### Geom_smooth

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth()

# Smoothed line shows dominant pattern and level of uncertainty
# Method = "loess" is default for small n (>1,000 points)

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth(span = 0.2)

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth(span = 1)

# Span control "wiggliness of line

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth(method = "lm")

# Linear model gives the line of best fit
```

#### Histograms and frequency polygons

```{r}
ggplot(mpg, aes(hwy)) + 
  geom_histogram()

ggplot(mpg, aes(hwy)) + 
  geom_freqpoly()

# Binwidth controls the width of bins (probably NOT ideal for your data, try experimenting)

ggplot(mpg, aes(hwy)) + 
  geom_freqpoly(binwidth = 1)

ggplot(mpg, aes(hwy)) + 
  geom_freqpoly(binwidth = 2.5)

p3 <- ggplot(mpg, aes(hwy)) + 
  geom_freqpoly(binwidth = 1)

p4 <- ggplot(mpg, aes(hwy)) + 
  geom_freqpoly(binwidth = 2.5)

grid.arrange(p3, p4, ncol = 2)

```

#### Bar chart

```{r}
ggplot(mpg, aes(manufacturer)) + 
  geom_bar()
```

#### Boxplot

```{r}
ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot()

ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot() +
  geom_point()

ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot() +
  geom_jitter()
```

#### Size and transparency control

```{r}
ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot() +
  geom_jitter(alpha = 0.2)

ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3)

ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, size = 2.5)

ggplot(mpg, aes(drv, hwy)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, size = 2.5, width = 0.1)
```

#### Modification of axes

```{r}
ggplot(mpg, aes(drv, hwy)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.25)

ggplot(mpg, aes(drv, hwy)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.25) + 
  xlim("f", "r") 

ggplot(mpg, aes(drv, hwy)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.25) + 
  xlim("f", "r")+ 
  ylim(20, 30)
#> Warning: Removed 138 rows containing missing values (`geom_point()`).

```

#### Title and axis names

```{r}
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5)

# Change axis titles
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)")

ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)")

# Remove the axis labels with NULL
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab(NULL) + 
  ylab(NULL) +
  ggtitle("City by Highway driving (mpg)")
```

#### Editing title and axis using Theme

```{r}
# Title
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

# Legend
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "left")

# Axis 
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"))

# Axis name
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

```

#### Overall theme

```{r}
# Theme black and white
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

# Theme position matters!
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold")) +
  theme_bw()

# Theme classic
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

# Theme minimal
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

# Theme dark
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme_dark() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

# Theme light
ggplot(mpg, aes(cty, hwy, color = class)) +
  geom_jitter(alpha = 0.3, size = 2.5) + 
  xlab("City driving (mpg)") + 
  ylab("Highway driving (mpg)") +
  ggtitle("City by Highway driving (mpg)") +
  theme_light() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), 
    legend.position = "right",
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

```

### Appendix

```{r show-code, ref.label = all_labels(), echo = TRUE, eval = FALSE}

```
