---
title: ""
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(flexdashboard)
```

### Instructions

The aim of this assignment is to provide you with an opportunity to sharpen your skills in using ggplot(). While some of the plots you create may resemble those we covered in class, it is essential that they are original.

Rules for the Assignment:

1.  All data sets except mpg are fair game. That includes past data sets used in the class, sets built into R, your own data, and even online data.

2.  All graphs must include axis labels, plot title, a theme of your choice, and a brief description/interpretation of the plot (2-3 sentences).

It's worth noting that ggplot() is a widely-used tool, and there are numerous online resources available for you to explore. We strongly recommend that you take advantage of these resources to deepen your understanding of ggplot().

Knit the document into a PDF and submit it to Canvas by 3/5/25 at 11:59 pm.

```{r echo = FALSE}
library(knitr)
library(tidyverse)
library(ggplot2)
library(ggthemes)
```

## Day 1 Assignment

### 0. Read in data

```{r}
births <- read_csv("US_births_2000-2014_SSA.csv", show_col_types = FALSE)

births$year <- as.factor(births$year)
births$month <- as.factor(births$month)
births$day_of_week <- as.factor(births$day_of_week)
```

### 1. Make a violin plot showing births by year (2pts)

```{r}

```

Description and Interpretation: What is the distribution of this data? What could drive this pattern of births across each year?

------------------------------------------------------------------------

### 2. Make a box plot showing births by month in 2000, include the jittered raw data (transparency to 0.5 and size to 2) (2pts).

```{r}

```

Description and Interpretation: Does the distribution of data change when looking across months versus across years?

------------------------------------------------------------------------

### 3. Make a plot showing births by day of week in January 2000 (2pts).

```{r}

```

Description and Interpretation: How does birth timing differ throughout the week?

------------------------------------------------------------------------

### 4. Based on newly generated columns, make a box plot (including raw data) of births by weekdays vs weekend days (2pts).

```{r}
births <- births %>%
  mutate(day_of_week_names = fct_recode(day_of_week,
                                  "Sunday" = "7",
                                  "Monday" = "1",
                                  "Tuesday" = "2",
                                  "Wednesday" = "3",
                                  "Thursday" = "4",
                                  "Friday" = "5",
                                  "Saturday" = "6")) %>%
  mutate(day_type = ifelse(day_of_week_names %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))


## Enter plot data below

```

Description and Interpretation: How does this plot differ from the in Question 3? What does it show?

------------------------------------------------------------------------

### 5. Run a statistical test to compare means between weekday and weekend (2pts).

```{r}

```

Description and Interpretation: What is the p-value of the statistical test? Does this indicate a significant difference in means between groups (weekday vs weekend)?

------------------------------------------------------------------------

## Day 2 Assignment

It's now time to try generating some more original plots! Please keep the assignment instructions in mind as you move through questions 6-10.

### Read in data

```{r}

```

### 6. Make a scatter plot with customized point size and transparency (3pts)

```{r}

```

Description and Interpretation:

------------------------------------------------------------------------

### 7. Fit a line through your scatter plot from 6. (3pts)

```{r}

```

Description and Interpretation:

------------------------------------------------------------------------

### 8. Make a new plot (different from above scatter plot) with facet_wrap() (3pts)

```{r}

```

Description and Interpretation:

------------------------------------------------------------------------

### 9. Using the plot from 8, change the colors and reorder your facet to reverse order. (3pts)

```{r}

```

Description and Interpretation:

------------------------------------------------------------------------

### 10. Make a Plot Using a Geometry That Was Not Covered in Class (3pts)

```{r}

```

Description and Interpretation:

### Appendix

```{r show-code, ref.label = all_labels(), echo = TRUE, eval = FALSE}
```
