---
title: "Bio File Formats in R"
author: "David C. King"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## File formats in biology

Many file formats in biology are plain text, but conform to a certain
structure.

-   Sequences- DNA, RNA, protein

-   Sequence reads - fastq

-   Sequence Alignments- MAF, SAM/BAM

-   Phylogenetic trees - Newick, Phylip

-   Genomic Coordinates and Annotations - BED, GTF, GFF

-   Microscopy images - tiff, raw

-   Protein Structure - pdb

-   Mass Spectrography

-   Analysis output – tidy data, csv

-   Compressed files - gzip, tar, zip

## DNA/RNA and Amino Acid Sequences

Purpose:

-   Understand sequence content, properties.

-   Domains/Structure/Function.

-   Variation: SNPs, indels, transitions/transversions.

-   Identity of a gene, relationship to other genes, other species
    (paralogs/orthologs).

-   Next-gen sequencing- the presence and identity of the sequence is a
    measurement (a count of expressed transcripts, enriched fragments)

-   Design primers for PCR

### FASTA Format

Comes from FAST sequence aligner. Fast**n**: nucleotide; Fast**p**:
protein: Fast**a**: all.

Definition:

<https://en.wikipedia.org/wiki/FASTA_format>

<https://www.ncbi.nlm.nih.gov/genbank/fastaformat/>

1.  Header line starts with "\>" followed by the name of the sequence.
    More information can be provided but must be on the same line.
2.  Sequence follows on one or more lines.
3.  Multiple header/sequence pairs can be in the same file (sometimes
    this is called "multi-fasta")

Examples:

```         
>gi|1817694395|ref|NZ_JAAGMU010000151.1| Streptomyces sp. SID7958 contig-52000002, whole genome shotgun sequence
CCGGCTGGCGCGGCTGGCGCTGGCGGTGGGGCTGCGGCTGCTGGAGCTGGGGGTGGCGCTGGAGGCGCAC
GGCCAGAACCTGCTGGTGGTGCTGTCGCCGTCCGGGGAGCCGCGGCGGCTGGTCTACCGCGATCTGGCGG
ACATCCGGGTCTCCCCCGCGCGGCTGGCCCGGCACGGTATCCGGGTTCCGGACCTGCCGGCG

>gi|1643051563|gb|SZWM01000399.1| Citrobacter sp. TBCS-14 contig3128, whole genome shotgun sequence
GCACAGTGAGATCAGCATTCCGTTGGATCTACTGGTCAATCAAAACCTGACGCTGGGTACTGAATGGAAC
CAGCAGCGCATGAAGGACATGCTGTCTAACTCGCAGACCTTTATGGGCGGTAATATTCCAGGCTACAGCA
GCACCGATCGCAGCCCATATTCGAAAGCCGAGATCTTCTCTTTGTTTGCCGAAAACAACATG
```

These sequences have a complex header with multiple, \|-delimited
fields.

What's missing? Sequence annotations and features.

-   Example: Say there is a stem-loop at position 150-200. Have to
    record it separately.

-   Exception: masking with uppercase/lowercase, or another symbol (i.e.
    N, X)

#### Masking

Sequence aligners are often tripped up by short repeat fragments and low
complexity regions. RepeatMasker is a common tool for finding repeats
and *masking* them.

One convention is to use the character N for a hard-masked DNA sequence,
or use lowercase letters for softmasking. *This doesn't work if N is
being used for an ambiguity (the exact base at that position is
uncertain).*

##### Hard Masking

(David made this example up)

In the Streptomyces example above, replace nucleotide letter with 'N' if
sequence is in a repeat (found by RepeatMasker)

```         
CCGGCTGGCGCGGCTGGCGCTGGCGGTGGGGCTGCGGCTGCTGGAGCTGGGGGNNNNNNNNNNNNNNNNN
NNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNNCTGGTCTACCGCGATCTGGCGG
ACATCCGGGTCTCCCCCGCGCGGCTGGCCCGGCACGGTATCCGGGTTCCGGACCTGCCGGCG
```

Software that deals with sequence analysis will be forced to ignore the
N's., but maintain the spacing.

##### Soft Masking

Sometimes you want to retain the sequence, but convey that it is masked.
This is usually done with the masked bases being converted to lowercase.

```         
CCGGCTGGCGCGGCTGGCGCTGGCGGTGGGGCTGCGGCTGCTGGAGCTGGGGGtggcgctggaggcgcac
ggccagaacctgctggtggtgctgtcgccgtccggggagccgcggcggCTGGTCTACCGCGATCTGGCGG
ACATCCGGGTCTCCCCCGCGCGGCTGGCCCGGCACGGTATCCGGGTTCCGGACCTGCCGGCG
```

Software might not take case into account, it depends on the
application, and behavior can be configured.

### In R

```{r fasta, include=FALSE}
# include=FALSE above because there are a lot of info messages when loading this library (clutters output)
library(Biostrings)
```

```{r fasta read in}

# Read from a fasta file
seqs = readDNAStringSet("seqs_for_markdown.fasta")
seqs

# Save the names because they are long, and use shorter ones
long_names = names(seqs)
names(seqs) = c("Streptomyces","Citrobacter")
seqs
```

```{r}
# Some functions

## Dinucleotide frequencies
oligonucleotideFrequency(seqs[1], 2)
oligonucleotideFrequency(seqs[2], 1, as.prob=T)


# look at hexamers for the most common ones
hexamer_frequencies = oligonucleotideFrequency(seqs[1], 6) # matrix of one row, 4096 columns
hexamer_frequencies[1,1:10] # can't use head on a wide object like this
descending_order = order(hexamer_frequencies, decreasing = TRUE) # returns the INDEX of the the largest hex frequency, followed by the next largest, etc.
descending_order[1:10] # top 10 hexamer indices
hexamer_frequencies[,descending_order[1:10]] # top ten hexamers (in a matrix)
top10 = names(hexamer_frequencies[,descending_order[1:10]]) # get the hexamers from the names


# turn the top10 into a sequence object
dict0 = DNAStringSet(top10)
pdict0 = PDict(dict0) # PDict for Pattern DICTionary

# find the occurences of the top 10 hexamers in the sequence
matches = matchPDict(pdict0, seqs[[1]])
for (i in 1:10) {print(top10[i]); print(replaceAt(seqs[[1]], matches[[i]], value = 'NNNNNN'))}
```

```{r Biostrings example}
expectedDinucleotideFrequency <- function(x)
{
    # Individual base frequencies.
    bf <- alphabetFrequency(x, baseOnly=TRUE)[DNA_BASES]
    (as.matrix(bf) %*% t(bf) - diag(bf)) / length(x)
}

## On Celegans chrI:
library(BSgenome.Celegans.UCSC.ce11)
chrI <- Celegans$chrI
obs_df <- dinucleotideFrequency(chrI, as.matrix=TRUE)
obs_df  # CG has the lowest frequency
exp_df <- expectedDinucleotideFrequency(chrI)
## A sanity check:
stopifnot(as.integer(sum(exp_df)) == sum(obs_df))

## Ratio of observed frequency to expected frequency:
obs_df / exp_df  # TA has the lowest ratio, not CG!
```

## Annotation Databases

Gene annotations, sequences, other data that can be mapped to genes from
a variety of sources/experiments.

NCBI - National Center for Biotechnology Information

UCSC Genome Browser - University of California Santa Cruz

[Ensembl](https://useast.ensembl.org/index.html "Ensembl Genome Browser") -
European Bioinformatics Institute

### Ensembl Genome Browser

[Ensembl *Canis lupus familiaris* HOXA4 (for German
Shephard)](https://useast.ensembl.org/Canis_lupus_familiarisgermanshepherd/Transcript/Exons?db=core;g=ENSCAFG00805011771;r=14:40378329-40397144;t=ENSCAFT00805021651 "Link to transcript view"){.uri}

[![HOX A4, German
Shephard](img/doghox_screenshot.png)](https://useast.ensembl.org/Canis_lupus_familiarisgermanshepherd/Transcript/Exons?db=core;g=ENSCAFG00805011771;r=14:40378329-40397144;t=ENSCAFT00805021651)

#### Tracks and Configuration

-   Tracks displayed in screenshot

    -   Contigs - sequence assembly
    -   Genes (Ensembl) - The Enbsembl gene annotations; others may be
        included

-   Configuration

    -   Tabs: Gene/Transcript
    -   Add/remove tracks
    -   Legend, Strand, scale
    -   Export options

##### Adding tracks

Click on *Region in Detail*

![HOX A4, German
Shephard](img/doghox_screenshot_highlight.png){width="66%"}

Chose some tracks to show, exported the image.

![HOX A4, German
Shephard](img/Dog___German_Shepherd_1440418485_40534882.png)

Changes:

- Zoomed out
- TSS (eponine) - Transciption start sites
- GC% Continuous, numeric data

Experiment with adding/hiding tracks:


![HOX A4, German
Shephard](img/Genome_view_highlighted.png)

##### For Comparison HOXA4 at UCSC

<https://genome.ucsc.edu/s/davidcking/canFam4_hoxA10>

They don't have the rest of the gene cluster!?!?!

### Exercise as a group

Purpose: brainstorm how to represent data in a dataframe

**What would be a good tidy way to represent track data in a dataframe?**

- What would the rows be?
- What would the columns be?
- How many observations per row?

