---
title: "Programming assignment"
author: "David C. King"
date: "`r Sys.Date()`"
output: html_document
---


## 1: Function Fundamentals

**1. `microscope_magnification()`**

Write a function called microscope_magnification() that calculates total magnification based on the objective and ocular lens magnification. Default ocular is 10x.

magnification = objective * ocular

### ✅ Solution

```{r my solution to functions Q1}

```


## Peak Percentages

I have determined a number of ChIP-seq peaks in putative enhancer
regions and would like to know what percentage of these peaks are
located on each chromosome. I have created code to calculate this, but
it is becoming extremely tedious to manually type each line...
especially since the yellow fever mosquito has over twenty thousand
unassembled contigs! Can you modify my code below so that it uses a loop
to print out the percentages for each unique chromosome in my dataframe?

```{r my solution to loops Peak Percentages}
enhancer_peaks=read.csv("LoopsConditionalsExerciseData/ExamplePeaksSheet.csv")

unique(enhancer_peaks$chromosome) # this gets the unique chromosomes (hint: save to a variable)

# Manual approach
print(sum(enhancer_peaks$chromosome=="AaegL5_1")/nrow(enhancer_peaks))
print(sum(enhancer_peaks$chromosome=="AaegL5_2")/nrow(enhancer_peaks))
print(sum(enhancer_peaks$chromosome=="AaegL5_3")/nrow(enhancer_peaks))
print(sum(enhancer_peaks$chromosome=="NIGP01002131")/nrow(enhancer_peaks))

# Loop approach
#for (#fill in#){
  # only need one command now:
  
  ####
#}
```